---
title: "Class 11 Candy"
author: "David Ma"
format: pdf
---


In today's class we will examine 558 candy data and see if this helps us gain some more feeling for how PCA and other methods work.

```{r}
candy <- read.csv("candy-data.csv", row.names = 1)
head(candy)

```

> Q1. How many different candy types are in this dataset?

```{r}
nrow(candy)
```

> Q2. How many fruity candy types are in this dataset?

```{r}
sum(candy$fruity)
```

> Q. What are these fruity candies?

```{r}
candy[ candy$fruity == 1, ]
```

## How often does my favorite candy win?

> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

```{r}
candy["Snickers", ]$winpercent
```

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy["Kit Kat", ]$winpercent
```

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```

```{r}
skimr::skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

Yes, the `winpercent` column is on a 0:100 scale as compared to others with a 0:1 scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A zero would represent that the candy does not contain chocolate while a one would represent that the candy contains chocolate.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy, aes(x = winpercent)) +
  geom_histogram(bins = 15)
```

> Q9. Is the distribution of winpercent values symmetrical?

No

> Q10. Is the center of the distribution above or below 50%

Below 50% with a mean

```{r}
mean(candy$winpercent)
```

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

To answer this question I will need to:

- "Subset" the candy dataset to just chocolate candy,
- get their winpercent values, 
- and then calculate the mean of these. 

Then do the same for fruity candy and compare. 

```{r}
chocolate_avg <- mean(candy$winpercent[as.logical(candy$chocolate)])
fruity_avg <-  mean(candy$winpercent[as.logical(candy$fruity)])

print(chocolate_avg)
print(fruity_avg)
```

On average, chocolate candy is ranked higher than fruity candy.

> Q12. Is this difference statistically significant?

```{r}
t.test((candy$winpercent[as.logical(candy$chocolate)]), candy$winpercent[as.logical(candy$fruity)])
```

## Overall Candy Rankings


There is a base R function called sort() for sorting vectors of input.

```{r}
x <- c(5,2,10)

# sort(x, decreasing = TRUE)
sort(x)
```

The buddy function to `sort()` that is often more useful is called `order()`. It returns the "indices" of the input that would result in it being sorted.

```{r}
order(x)
```

> Q13. What are the five least liked candy types in this set?

```{r}
library(tidyverse)
candy %>%
  arrange(winpercent) %>%
  head(5)
```

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy %>%
  arrange(desc(winpercent)) %>%
  head(5)
```

> Q15. Make a barplot of candy ranking using `winpercent` values.

```{r}
ggplot(candy) +
  aes(winpercent, rownames(candy)) +
  geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy) +
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col()
```

> Adding colors to these bars

```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```

```{r}
ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols) 
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starburst

## Taking a look at pricepoint

> Q. What is the best candy for the least money?

```{r}
library(ggrepel)

# How about a plot of price vs win
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  # ggrepel helps deal with overlapping labels
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 30)
```

> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

Reese's miniatures

> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```
Nik L Nip

## Exploring the correlation structure

Pearson correlation goes between -` and +1 with zero indicating no correlation. Values close to one indicate being very highly correlated.

```{r}
library(corrplot)
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Chocolate and fruit are ani-correlated.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and winpercent or chocolate and bar.

## Principal Component Analysis

The base r function for PCA is called `prcomp()` and we can set "scale=TRUE/FALSE".

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

The main result of PCA - i.e. the new PC plot (projection of candy on our new PC axis) is contained in `pca$x`

```{r}
pc <- as.data.frame(pca$x)

ggplot(pc) +
  aes(PC1, PC2, label = rownames(pc)) +
  geom_point(col = my_cols) +
  geom_text_repel(max.overlaps  = 30)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 30)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
#library(plotly)
#ggplotly(p)

```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

Fruity, hard, and pluribus are picked up strongly by PC1 in the positive direction. These make sense these variables are strongly correlated, as seen in our correlation structure.
